---
title: "Getting started with census2020download"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with census2020download}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # The download examples reach the Census Bureau over the network, so they are
  # shown but not executed when the vignette is built.
  eval = FALSE
)
```

## What this package does

`census2020download` is a small set of helper functions used by the
[EJAM](https://ejanalysis.com/ejamdocs) package. It downloads the 2020 Census
redistricting (PL94-171) files for US states, DC, and Puerto Rico — plus the
Demographic and Housing Characteristics (DHC) files for the Island Areas
(VI, GU, MP, AS) — then unzips, reads, and reshapes them into compact
`data.table`s.

The full pipeline has four steps, wrapped by the single function `census2020_get_data()`:

1. `census2020_download()` — fetch the zip files from the Census Bureau
2. `census2020_unzip()` — unzip the data and geographic header files
3. `census2020_read()` — read and merge the pipe-delimited files
4. `census2020_clean()` — rename columns, compute total area, subset

```{r}
library(census2020download)
```

## The one-call pipeline

`census2020_get_data()` runs all four steps. Try it with a couple of small
states so the download is quick:

```{r}
blocks <- census2020_get_data(c("DE", "DC"))
dim(blocks)
head(blocks)
```

By default it returns the columns `blockfips`, `lat`, `lon`, `pop`, and `area`
(area in square meters). To keep every available column — including the
race/ethnicity breakdown — pass `cols_to_keep = "all"`:

```{r}
blocks_all <- census2020_get_data(c("DE", "DC"), cols_to_keep = "all")

# The race/ethnicity subgroups sum exactly to the total block population:
groups <- c("hisp", "nhwa", "nhba", "nhaiana", "nhaa",
            "nhnhpia", "nhotheralone", "nhmulti")
all.equal(blocks_all$pop, rowSums(blocks_all[, ..groups]))
```

## Controlling downloads

* `folder` — where zip files are saved (default: a temporary directory).
* `overwrite = FALSE` — skip downloading any zip already present in `folder`.
* `timeout` — seconds before an individual download times out (default 180).

```{r}
blocks <- census2020_get_data(
  c("DE", "DC"),
  folder = "~/census2020zip",
  overwrite = FALSE,
  timeout = 300
)
```

## Building the EJAM data tables

`census2020_save_datasets()` splits the cleaned blocks into the five tables
EJAM uses, and computes each block's population weight within its parent
block group:

```{r}
tables <- census2020_save_datasets(blocks)
names(tables)
#> "bgid2fips" "blockid2fips" "blockpoints" "blockwts" "quaddata"
```

## Island Areas

Block-level data are not published for the Island Areas, so these come at
block-group resolution via a dedicated helper:

```{r}
islands <- census2020_get_data_islandareas()  # VI, GU, MP, AS
```

`census2020_get_data()` can also accept a mix of mainland states and Island
Areas and will dispatch each to the appropriate reader.

## Technical references

See the help for `?census2020_read` for links to the Census Bureau technical
documentation describing the files, tables, and variables.
